---
title: "30 Year Climate Trends"
author: "Megan Hoover"
format: html
editor: visual
---

# 30 Year Climate Trends in CO, NE, SD, and WY <br>
Organization: CNHP
Affiliates: Megan Hoover, Andrew Zimbelman, Chippy Marx

We downloaded monthly and daily climate data from https://prism.oregonstate.edu/ based off of the species and locations we provided. We are comparing 1990-2020 monthly precipitation and temperature trends to 2026 trends. 

```{r}
#| include: false #prevents code and output from showing
library(sf)
library(tidyverse)
library(ggplot2)
library(leaflet)
library(plotly)
library(gganimate)
library(leaflet)
library(tmap)
library(FNN)
library(shiny)
```
 
```{r}
#| include: false
setwd("C:/Users/samho/Desktop/CNHP/Growing_Degree_days/GDD_CNHP")
#Load prism climate data csv files
m1<- read_csv("raw_data/PRISM_ppt_tmean_monthly_199101_200501.csv")
m1 <- m1 %>%
  rename(tmean_F = `tmean_ F`)
m2<- read_csv(
  "raw_data/PRISM_ppt_tmean_monthly200502_202001.csv")
m2 <- m2 %>% filter(!is.na("tmean_F")) %>% 
  filter(!is.na(ppt_in))
m1 <- m1 %>% filter(!is.na("tmean_F")) %>% 
  filter(!is.na(ppt_in))
m3<-read_csv("raw_data/all_prismsp_monthly_temp_precip_2026v2.csv") %>%
  rename(species=Name)

#bind all data into one
dat<-bind_rows(m1,m2,m3)

```
 

```{r}
#| include: false
#changing date format and making seasons
dat <- dat %>%
  mutate(
    Date = lubridate::ym(Date),
    year = lubridate::year(Date),
    month = lubridate::month(Date))#,
   # season = case_when(
   # month %in% c(12, 1, 2) ~ "winter",
   # month %in% c(3, 4, 5) ~ "spring",
   # month %in% c(6, 7, 8) ~ "summer",
   # month %in% c(9, 10, 11) ~ "fall")) 

#grouping by species, location, and year to compare trends
dat1<- dat %>%
  group_by(species, Latitude, Longitude, month,year) %>%
  summarise(
    temp = mean(tmean_F, na.rm = TRUE),
    precip = sum(ppt_in, na.rm = TRUE),
    .groups = "drop") %>%
  mutate(month = as.integer(month))

#trying anomlies instead of raw temps. shows difference b/w raw temp and mean temperature per month.
dat_anom <- dat1 %>%
  group_by(species, Latitude, Longitude, month) %>%
  mutate(temp_anom = temp - mean(temp, na.rm = TRUE)) %>%
  ungroup()
  
tmap_mode("view")
```


**Pixelated heat map of the 30 year temperature trends compared to 2026**

There are only climate data on 6 of the 8 plant species we were interested in.
```{r}
#| include: false
#collapsed species data for avg temp per species per month per year. Can do this b/c same eco region for all species. This is more simple and easier to plot
dat_region <- dat1 %>%
  group_by(species, month, year) %>%
  summarise(
    temp = mean(temp, na.rm = TRUE),
    .groups = "drop")
```


```{r}
ggplot(dat_region, aes(year, factor(month), fill = temp)) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_viridis_c(option="inferno") +
  theme_minimal() +
  labs(
    title = "Mean Monthly Temperature Trends 1990-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Temp(°F)")
```



**Pixelated precipitation map of the 30 year trends compared to 2026**
```{r}
#| include: false
dat_region <- dat1 %>%
  group_by(species, month, year) %>%
  summarise(
    precip = mean(precip, na.rm = TRUE),
    .groups = "drop")
```

```{r}
ggplot(dat_region, aes(year, factor(month), fill = precip)) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_viridis_c(option= "turbo") +
  theme_minimal() +
  labs(
    title = "Mean Monthly Precipitation Trends 1990-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Precip(in)")
  
```




```{r}
#building a shiny ap that is interactive
ui <- fluidPage(
  leafletOutput("map", height = 600),
  plotOutput("plot", height = 300)
)

server <- function(input, output, session) {

  # ---------------- MAP ----------------
  output$map <- renderLeaflet({

    leaflet(dat) %>%
      addTiles() %>%

      # 🌱 species points (clickable)
      addCircleMarkers(
        lng = ~Longitude,
        lat = ~Latitude,
        radius = 5,
        color = "darkgreen",
        fillOpacity = 0.8,
        layerId = ~species   # assumes species column exists
      )
  })

  # ---------------- CLICK ----------------
  observeEvent(input$map_marker_click, {

    click <- input$map_marker_click

    # 🌡 nearest climate pixel (same df!)
    idx <- get.knnx(
      dat[, c("Longitude", "Latitude")],
      matrix(c(click$lng, click$lat), ncol = 2),
      k = 1
    )$nn.index

    df <- dat[idx, ]

    # collapse to time series
    ts <- df %>%
      group_by(year, season) %>%
      summarise(
        tmean = mean(tmean_F, na.rm = TRUE),
        ppt   = mean(ppt_in, na.rm = TRUE),
        gdd   = mean(GDD, na.rm = TRUE),
        .groups = "drop"
      )

    # ---------------- PLOT ----------------
    output$plot <- renderPlot({

      ggplot(ts, aes(x = year)) +

        geom_col(aes(y = tmean, fill = "Temp"), alpha = 0.6) +
        geom_line(aes(y = ppt * 10, color = "Precip")) +
        geom_line(aes(y = gdd, color = "GDD")) +

        facet_wrap(~season) +

        scale_fill_manual(values = c("Temp" = "firebrick")) +
        scale_color_manual(values = c("Precip" = "steelblue",
                                      "GDD" = "darkgreen")) +

        theme_minimal() +
        labs(
          title = paste("Climate Trend:", click$id),
          x = "Year",
          y = "Value (scaled)"
        )
    })

  })
}

shinyApp(ui, server)
```




The `echo: false` option disables the printing of code (only output is displayed).
