---
title: "30 Year Climate Trends"
author: "Megan Hoover"
format: html
editor: visual
---

# 30 Year Climate Trends in CO, NE, SD, and WY <br>
Organization: CNHP
Affiliates: Megan Hoover, Andrew Zimbelman, Chippy Marx

We downloaded monthly and daily climate data from https://prism.oregonstate.edu/ based off of the species and locations we provided. We are comparing 1990-2020 monthly precipitation and temperature trends to 2026 trends. 

```{r}
#| include: false #prevents code and output from showing
library(sf)
library(tidyverse)
library(ggplot2)
library(leaflet)
library(plotly)
library(gganimate)
library(leaflet)
library(tmap)
library(FNN)
library(shiny)
```
 
```{r}
#| include: false
setwd("C:/Users/samho/Desktop/CNHP/Growing_Degree_days/GDD_CNHP")
#Load prism climate data csv files
m1<- read_csv("raw_data/PRISM_ppt_tmean_monthly_199101_200501v3.csv")
m1 <- m1 %>%
  rename(tmean_F = `tmean_ F`)
m2<- read_csv(
  "raw_data/PRISM_ppt_tmean_monthly200502_202001v3.csv")
m2 <- m2 %>% filter(!is.na("tmean_F")) %>% 
  filter(!is.na(ppt_in))
m1 <- m1 %>% filter(!is.na("tmean_F")) %>% 
  filter(!is.na(ppt_in))
m3<-read_csv("raw_data/all_prismsp_monthly_temp_precip_2026v3.csv") %>%
  rename(species=Name,
         Elevation_ft=`Elevation (ft)`,
         ppt_in= `ppt (inches)`,
         tmean_F=`tmean (degrees F)`) %>%
  filter(!is.na(ppt_in))

#bind all data into one
dat<-bind_rows(m1,m2,m3) %>%
  filter(!is.na(species))
```
 

```{r}
#| include: false
#changing date format and making seasons
dat <- dat %>%
  mutate(
    Date = lubridate::ym(Date),
    year = lubridate::year(Date),
    month = lubridate::month(Date))#,
   # season = case_when(
   # month %in% c(12, 1, 2) ~ "winter",
   # month %in% c(3, 4, 5) ~ "spring",
   # month %in% c(6, 7, 8) ~ "summer",
   # month %in% c(9, 10, 11) ~ "fall")) 

#grouping by species, location, and year to compare trends
dat1<- dat %>%
  group_by(species, Latitude, Longitude, month,year) %>%
 # summarise(
  #  temp = mean(tmean_F, na.rm = TRUE),
  #  precip = sum(ppt_in, na.rm = TRUE),
  #  .groups = "drop") %>%
  mutate(month = as.integer(month))

#trying anomlies instead of raw temps. shows difference b/w raw temp and mean temperature per month.

#Jan. values are compared to the average January (across all 30 years, etc.
dat_anom <- dat1 %>%
  group_by(species, Latitude, Longitude, month) %>%
  mutate(
    temp_climatology = mean(tmean_F, na.rm = TRUE),
    temp_anom = tmean_F - temp_climatology
  ) %>%
  ungroup()
  
tmap_mode("view")
```


**Pixelated heat map of the 30 year temperature trends compared to 2026**

```{r}
#| include: false
#collapsed species data for avg temp per species per month per year. Can do this b/c same eco region for all species. This is more simple and easier to plot
dat_region <- dat1 %>%
  group_by(species, month, year) #%>%
 # summarise(
  # temp = mean(tmean_F, na.rm = TRUE),
  #  .groups = "drop")
```


```{r}
ggplot(dat_region, aes(year, factor(month), fill = tmean_F)) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_viridis_c(option="inferno") +
  theme_minimal() +
  labs(
    title = "Mean Monthly Temperature Trends 1990-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Temp(°F)")
```

**Pixelated heat map of the temperature anomaly across species from 1991-2020**

Anomalies are compared to a baseline value and they remove seasonality so you can compare across time more meaningfully.
```{r}
#plot temp anomaly. Good to see for changes across years
ggplot(dat_anom, aes(x = year, y = factor(month), fill = temp_anom)) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +labs(
    title = "Mean Monthly Temperature Anonomly Trends 1991-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Temperature Change(F)")
  
```


**Pixelated precipitation map of the 30 year trends compared to 2026**
```{r}
#| include: false
dat_region <- dat1 %>%
  group_by(species, month, year) #%>%
 # summarise(
    #precip = mean(precip, na.rm = TRUE),
   # .groups = "drop")

#precip anomaly
dat_anom <- dat1 %>%
  group_by(species, Latitude, Longitude, month) %>%
  mutate(
    p_climatology = mean(ppt_in, na.rm = TRUE),
    p_anom = ppt_in - p_climatology
  ) %>%
  ungroup()
```

```{r}
ggplot(dat_region, aes(year, factor(month), fill = ppt_in)) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_viridis_c(option="mako", direction = -1) + #direction = -1 reverse palette
  theme_minimal() +
  labs(
    title = "Mean Monthly Precipitation Trends 1990-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Precip(in)")
  
```
**Pixelated Precipitation anomaly map of the plant species from 1991-2020**

```{r}
#plot temp anomaly. Good to see for changes across years
ggplot(dat_anom, aes(x = year, y = factor(month), fill = p_anom)) +geom_tile() +
  facet_wrap(~ species) +
     scale_fill_gradient2(low = "red", mid = "white", high = "blue") + theme_minimal() +
  labs(
    title = "Mean Monthly Precipitation Anonomly Trends 1991-2020, 2026",
    x = "Year",
    y = "Month",
    fill = "Precipitation Change(in)")
```


Now here is a map that shows where all the locations of the plant that we would like to collect on in 2026. The GDD days are shown with each species, to show how much more heat energy the plant has left until ready for harvest.


```{r}
#| include: false
#Read in the GDD and plant location csv and make into sf
gdd<-read_csv("clean_data/latest_gdd_by_location2.csv") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
gdd<- gdd %>%
  rename(species=species_code,
         GDD=med_total_gdd)

#read in the states sf file from CO, NE, KS, WY
states<-st_read("raw_data/states.shp")
#check that they match
st_crs(states) == st_crs(gdd)
#project states to 4326
states<- st_transform(states, 4326)
```


**Interactive map of plant species of interest and their GDD. Only 6 species are shown**
```{r}
tmap_mode("view")  # for interactive map and static for non-interactive

#tm<-
  tm_shape(states)+
  tm_polygons(col="grey")+
tm_shape(gdd)+
  tm_dots(col = "species",
          size = 0.5,
          alpha = 1.0,
          popup.vars = c("species", "GDD"))
```





The `echo: false` option disables the printing of code (only output is displayed).
