---
title: "Growing Degree Days Model"
author: "Megan Hoover, Chippy, Andrew"
date: "2026-02-11"
output: html_document
---

# We will clean field data for plants found in the U.S. National Forest Service Area territory in 

Notes on common syntax within Rmd:
- If you want space between words or code chunks, write <br>
- If you want bold words/ headings use double stars **whateverwords**

*For Pushing work from Rstudio to repository in terminal: 5 steps*
- git pull
- save script
- git add .
- git commit -m "lil note to look back at or updates made to code"
- git push

*For Pulling work:*
- Click the Git tab, then click on pull

*Having conflict Merges?*
1. click on git tab, then select diff
2. Look for conflicts:
<<<<<<< HEAD
your version of text / code
=======

your teammate’s version of text / code
`>>>>>>> origin/main

3. Go back to git tab and leave the review changes you were just in from the diff. Right click on the rmd file that is staged, click open file
4. Delete those symbols in step 2. This allows for both changes from group members to stay, but it does it in order of who pushed first. 

**OR if needed to pull before commiting changes:**
git pull --rebase origin main
This will:

Download what's on GitHub

Place your new commit on top of it

Avoid a messy merge commit

Then push again:

git push
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

Load Libraries:
```{r}
library(tidyverse)
library(sf)
```
<br>

**Load field and spatial Data** example code given
```{r}
plants23<-read_csv("Data/name_of_file.csv") 
plants24<-read_csv("Data/name_of_file.csv") 

spatial<--read_csv("Data/name_of_file.csv")

#check structure/ what format data is in
str(name of df)

```


**Join the files**

inner join: Only columns with matching values in both data sets are kept in the result. May need to change column names.
```{r}
#example
plants_joined<- df1 %>% 
  inner_join(df2 %>% select(col1, col4)) #joining on named columns, can have several 
```

